---
title: "homework 2"
author: "Polly Wu (rw3031)"
date: "2024-09-24"
output: github_document
---

```{r setup, echo=FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

# Problem 1

### import and tidy the NYC transit dataset 

```{r}
subway = 
  read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv",  
           na = c("NA", ".",""))|>
  janitor::clean_names()|>
  select(line:entry, vending, ada)|>
  mutate(entry = recode(entry, "YES" = TRUE ,"NO" = FALSE))
```
### describe the dataset

The subway dataset contains variables including: `r names(subway)`.

For cleaning up the data, I convert the format of all of the column names into lower case and connected through underscore. I have retain the variables that we are interested in and convert the entry column into a logical variable. 

There are `r ncol(subway)` columns and `r nrow(subway)` rows. 

The data is not tidy since we still have the route column in wide format.Each route is stored in a different column instead of consolidated together.

```{r}
station = distinct(subway,line,station_name, .keep_all = TRUE)
nrow(station)
nrow(filter(station,ada == "TRUE"))
nrow(filter(station,vending == "NO"))/nrow(station)*100
```
There are 465 distinct stations.
84 stations are ADA compliant. 
1.94% of the stations entrance/exit are without vending allow entrance. 

### tidy up the route variable

```{r}
subway =
  subway|>
  mutate(route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11),)|>
  pivot_longer(
    route1:route11,
    names_to = "route_no",
    names_prefix = "route",
    values_to = "route_name"
  )
```

### A_train stations

```{r}
station = distinct(subway,line,station_name, .keep_all = TRUE)
nrow(filter(station,route_name == "A"))
nrow(filter(station,route_name =="A", ada == "TRUE"))
```
There are 60 distinct stations serve the A train and 17 of them are ADA compliance. 

# Problem 2

### import the trash-wheel dataset 
```{r}
mr_trash = 
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", skip = 1,  na = c("NA", ".","")) |>
  janitor::clean_names()|>
  select(dumpster, weight_tons:homes_powered)|>
  mutate(sports_balls=as.integer(sports_balls), type = "mr.")
```
```{r}
professor_trash = 
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1,  na = c("NA", ".","")) |>
  janitor::clean_names()|>
  select(dumpster, weight_tons:homes_powered)|>
  mutate(type = "professor")
```

```{r}
Gwynnda = 
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1,  na = c("NA", ".","")) |>
  janitor::clean_names()|>
  select(dumpster, weight_tons:homes_powered)|>
  mutate(type = "gwynnda")
```

### merge the three datasets 
```{r}
trash_wheel = 
  bind_rows(mr_trash, professor_trash, Gwynnda)
```

The merged trash_wheel dataset has the follow variables `r names(trash_wheel)`. There are `r nrow(trash_wheel)`observations and `r ncol(trash_wheel)` variables. `

# Problem 3
